---
title: "DRIMSeq DTU"
author: ""
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
    code_folding: show
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script performs differential transcript usage analysis with DRIMSeq, based
on abundance estimates from Salmon. It supports testing one or more contrasts.

# Load packages

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(tximport)
  library(tximeta)
  library(SummarizedExperiment)
  library(edgeR)
  library(DRIMSeq)
  library(ggplot2)
})
```

# Load `SummarizedExperiment` object

We will use the transcript-level quantifications.

```{r}
sg <- se$sg
st <- se$st
```

# Create dmDSdata object

```{r dmDSdata}
counts <- data.frame(feature_id = rownames(st),
                     gene_id = unlist(rowData(st)$gene_id),
                     assays(st)[["counts"]],
                     row.names = NULL)

metadata <- data.frame(colData(st))
metadata <- metadata %>% 
    dplyr::select(sample_id = names, group  = celline)
  
d <- dmDSdata(counts = counts, samples = metadata)
plotData(d)
```

# Filter ************** MODIFY ************** 

```{r data-filter}
d <- dmFilter(d, min_samps_gene_expr = 3, min_samps_feature_expr = 3,
              min_gene_expr = 10, min_feature_expr = 5)
plotData(d)
```

# Define design. ************** MODIFY ************** 

```{r define-design}
# (des <- model.matrix(~ XXXX, data = samples(d)))
(des <- model.matrix(~ 0 + group, data = samples(d)))
```

# Calculate precision

```{r calculate-precision}
set.seed(123)
d <- dmPrecision(d, design = des)
plotPrecision(d)
```

# Fit model

```{r fit-model}
d <- dmFit(d, design = des, verbose = 1)
```

# Define contrasts. ************** MODIFY ************** 

```{r define-contrasts}
#(contrasts <- as.data.frame(makeContrasts(XXXX, levels = des)))
(contrasts <- as.data.frame(makeContrasts("groupN61311-groupN052611", levels = des)))
```

# Perform tests ************** MODIFY **************

```{r result-genes}
level <- "gene" ## set to "feature" if transcript-level results are desired
signif3 <- function(x) signif(x, digits = 3)
DRIMSeq_fits <- lapply(contrasts, function(cm) {
  dr <- dmTest(d, contrast = cm, verbose = 1)
  print(plotPValues(dr, level = level))
  dr
})

DRIMSeq_res <- lapply(DRIMSeq_fits, function(dr) {
  results(dr, level = level) %>% dplyr::mutate_if(is.numeric, signif3)
})
```


```{r result-transcripts}
level <- "feature" ## set to "feature" if transcript-level results are desired
DRIMSeq_feature_fits <- lapply(contrasts, function(cm) {
  dr <- dmTest(d, contrast = cm, verbose = 1)
  print(plotPValues(dr, level = level))
  dr
})

DRIMSeq_feature_res <- lapply(DRIMSeq_feature_fits, function(dr) {
  results(dr, level = level) %>% dplyr::mutate_if(is.numeric, signif3)
})
```
# Write results to text files and save .rds file

```{r save-result}
if (class(DRIMSeq_res) == "data.frame") {
  write.table(DRIMSeq_res %>% dplyr::arrange(pvalue), 
              file = "DRIMSeq_dtu_results.txt", 
              sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
} else {
  for (nm in names(DRIMSeq_res)) {
    write.table(DRIMSeq_res[[nm]] %>% dplyr::arrange(pvalue), 
                file = paste0("DRIMSeq_dtu_results_", nm, ".txt"), 
                sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
  }
}

saveRDS(list(results = DRIMSeq_res, fits = DRIMSeq_fits), 
        file = "DRIMSeq_dtu.rds")
```


# Output results as `SummarizedExperiment` object

The `SummarizedExperiment` object on the transcript level
```{r se-gene}
## add rows (NA) for genes that are filtered out (if any)
DRIMSeq_resA <- lapply(seq_along(DRIMSeq_res), FUN = function(x) {
    
    # all genes
    geneA <- rowData(sg)$gene_id
    
    # genes that are not filtered out
    resX <- DRIMSeq_res[[x]]
    
    # other characteristics that have been calculated
    mexp <- mean_expression(DRIMSeq_fits[[x]]) %>%
        dplyr::arrange(match(gene_id, resX$gene_id)) %>%
        dplyr::select(-gene_id) 
    prec <- genewise_precision(DRIMSeq_fits[[x]]) %>%  
        dplyr::arrange(match(gene_id, resX$gene_id)) %>%
        dplyr::select(-gene_id) 
   
    resX <- resX %>%
        dplyr::bind_cols(mexp) %>%
        dplyr::bind_cols(prec) %>%
        dplyr::mutate(contrast = names(DRIMSeq_res)[[x]]) 
    resX$common_precision <- common_precision(DRIMSeq_fits[[x]])
    rownames(resX) <- resX$gene_id
    
    # genes that are filtered out 
    geneO <- setdiff(geneA, resX$gene_id)
    
    # results for all genes
    if (length(geneO) > 0) {
        # create a data frame with values NA as the results of the genes that
        # are filtered out
        matO <- matrix(NA, nrow = length(geneO),
                        ncol = ncol(resX),
                        dimnames = list(geneO, 
                                        colnames(resX))) 
        resO <- data.frame(matO)
        resO$gene_id <- geneO
        
        # combine the result tables 
        resA <- resO %>%
            dplyr::bind_rows(resX) %>%
            dplyr::arrange(match(gene_id, geneA)) %>%
            dplyr::mutate(contrast = names(DRIMSeq_res)[[x]]) 
            
    } else {
        resA <- resX %>%
            dplyr::arrange(match(gene_id, geneA)) %>%
            dplyr::mutate(contrast = names(DRIMSeq_res)[[x]]) 
             
    }
    
    # use gene column as rownames
    rownames(resA) <- resA$gene_id
    
    # convert to DataFrame
    resA <- S4Vectors::DataFrame(resA)
    return(resA)
})
names(DRIMSeq_resA) <- names(DRIMSeq_res)

## Put the result tables in rowData 
for (i in seq_along(DRIMSeq_resA)) {
    nam <- names(DRIMSeq_resA)[i]
    namI <- paste("DRIMSeq:", nam, sep = "")
    rowData(sg)[[namI]] <- DRIMSeq_resA[[i]] 
}


```


The `SummarizedExperiment` object on the transcript level
```{r se-tx}
## add rows (NA) for genes that are filtered out (if any)
DRIMSeq_resB <- lapply(seq_along(DRIMSeq_feature_res), FUN = function(x) {
    
    # all genes
    txA <- rowData(st)$tx_id
    
    # genes that are not filtered out
    resX <- DRIMSeq_feature_res[[x]]
    
    prop <- proportions(DRIMSeq_feature_fits[[x]]) %>%  
        dplyr::arrange(match(feature_id, resX$feature_id)) %>%
        dplyr::select(-c(gene_id, feature_id))
    colnames(prop) <- paste("proportion", colnames(prop), sep = "_")
    
    coef <- coefficients(DRIMSeq_feature_fits[[x]]) %>%  
        dplyr::arrange(match(feature_id, resX$feature_id)) %>%
        dplyr::select(-c(gene_id, feature_id))
    colnames(coef) <- paste("coef", colnames(coef), sep = "_")
    
    resX <- resX %>%
        dplyr::bind_cols(prop) %>%
        dplyr::bind_cols(coef) %>%
        dplyr::mutate(contrast = names(DRIMSeq_res)[[x]]) 
    
    rownames(resX) <- resX$feature_id
    
    # genes that are filtered out 
    txO <- setdiff(txA, resX$feature_id)
    
    # results for all genes
    if (length(txO) > 0) {
        # create a data frame with values NA as the results of the genes that
        # are filtered out
        matO <- matrix(NA, nrow = length(txO),
                        ncol = ncol(resX),
                        dimnames = list(txO, 
                                        colnames(resX))) 
        resO <- data.frame(matO)
        resO$feature_id <- txO
        
        # combine the result tables 
        resA <- resO %>%
            dplyr::bind_rows(resX) %>%
            dplyr::arrange(match(feature_id, txA)) %>%
            dplyr::mutate(contrast = names(DRIMSeq_feature_res)[[x]]) 
            
    } else {
        resA <- resX %>%
            dplyr::arrange(match(feature_id, txA)) %>%
            dplyr::mutate(contrast = names(DRIMSeq_feature_res)[[x]]) 
             
    }
    
    # use gene column as rownames
    rownames(resA) <- resA$feature_id
    
    # convert to DataFrame
    resA <- S4Vectors::DataFrame(resA)
    return(resA)
})
names(DRIMSeq_resB) <- names(DRIMSeq_feature_res)

## Put the result tables in rowData 
for (i in seq_along(DRIMSeq_resB)) {
    nam <- names(DRIMSeq_resB)[i]
    namI <- paste("DRIMSeq:", nam, sep = "")
    rowData(st)[[namI]] <- DRIMSeq_resB[[i]] 
}
```

The output is saved as a list.
```{r save-se}
analysis_se <- c(se, list(DRIMSeq_sg = sg, DRIMSeq_st = st))
saveRDS(analysis_se, file = "analysis_se.rds")
```
# Session info

```{r}
date()
sessionInfo()
```

