---
title: "Shiny preparation"
author: ""
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
    code_folding: show
    keep_md: true
editor_options: 
  chunk_output_type: console
---

```{r shiny-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "pdf"))
```

# Introduction

This script prepares the data to be used for shiny app. 

# Load packages

```{r shiny-load-pkg}
suppressPackageStartupMessages({
  library(tibble)
  library(dplyr)
  library(tidyr)
  library(limma)
  library(edgeR)
  library(reshape2)
  library(SingleCellExperiment)
  library(S4Vectors)
})
```

# Load data

```{r shiny-load-data}
options(ucscChromosomeNames = FALSE)
sg <- se$sg
st <- se$st
```

# Gene models

```{r shiny-gene-model}
create_genemodels <- function(genemodels) {
    idx <- match(c("transcript_id", "gene_id", "exon_id"), 
                 colnames(mcols(genemodels)))
    colnames(mcols(genemodels))[idx] <- c("transcript", "gene", "exon")
    mcols(genemodels)$symbol <- mcols(genemodels)$transcript
    subset(genemodels, type == "exon")
}

if (!is.null(gtffile)) {
    genemodels <- create_genemodels(genemodels)
} else {
    genemodels <- NULL
}
```

# Vector with bigWig file names 

```{r shiny-bigwig}
if (!is.null(bigwigdir)) {
    bwfiles <- normalizePath(list.files(bigwigdir, pattern = "\\.bw$", 
                                        full.names = TRUE))
    names(bwfiles) <- gsub("_Aligned.sortedByCoord.out.bw", "", basename(bwfiles))
} else {
    bwfiles <- NA
}
```

# edgeR - gene-level MDS

```{r shiny-MDS}
logcpms <- assay(sg, "logcpm")
mds <- limma::plotMDS(logcpms, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(logcpms),
                      MDS1 = mds$x,
                      MDS2 = mds$y)
}
mds <- mds %>% 
    dplyr::full_join(data.frame(colData(sg)), by = "names")
```

# SingleCellExperiment on gene level

The `rowData` of `sce_gene` includes the gene information and the result tables
from `edgeR` and `DRIMSeq`. Each result table is stored as a column, and the
column name is composed by `edgeR:` or `DRIMSeq:` and the name of the contrast
used. 

The `colData` of `sce_gene` stores the sample information, the bigWig file names
and condition information

The multidimensional scale data is stored in `reducedDims`.

```{r shiny-sce-gene}
nam <- colData(sg)$names

## low dimensional representation
reducedData <- mds %>%
    dplyr::arrange(match(names, nam)) %>%
    as.data.frame() %>% 
    dplyr::mutate(namestmp = names) %>%
    tibble::column_to_rownames("namestmp") %>%
    dplyr::select(-one_of(colnames(colData(sg))))
reducedData <- as.matrix(reducedData)

## column data
colData(sg)$bwFiles <- bwfiles[nam]

sce_gene <- SingleCellExperiment(assays = assays(sg), 
                                 rowData = rowData(sg),
                                 colData = colData(sg),
                                 metadata = list(geneModels = genemodels),
                                 reducedDims = SimpleList(MDS = reducedData))
```

# SingleCellExperiment on transcript level

The `rowData` of `sce_tx` includes the information of genes and transcripts,
and the result table on the transcript level from `DRIMSeq`.

The `colData` of `sce_tx` stores the sample information, the bigWig file names
and condition information.

```{r shiny-sce-tx}
nam <- colData(st)$names

## column data
colData(st)$bwFiles <- bwfiles[nam]

sce_tx <- SingleCellExperiment(assays = assays(st), 
                               rowData = rowData(st),
                               colData = colData(st),
                               metadata = list(geneModels = genemodels))
```

# Output results

```{r shiny-save-sce}
saveRDS(list(sce_tx = sce_tx, 
             sce_gene = sce_gene),
        file = "shiny_sce.rds")
```

# Session info

The analyses above were performed with the following package versions:

```{r shiny-session-info}
sessionInfo()
date()
```

